The language of generalization: Probability, vagueness, and interaction

### 2018-03-22 paper/case2-habituals.Rmd
small edits to habituals prose, redo hab priors figure

The event of Mary smoking at 2:30pm on January the 24th is an instance of a kind of event: *Mary smoking*. 
Just as with categories, generalization can occur over events of a particular type.
When expressed in language, these generalizations take the form of *habitual language*: "Mary smokes".

Events are a broad class of entities. 
In our second case study, we focus on habituals about people's behaviors that take the form: \textsc{singular noun phrase} $+$ \textsc{present tense simple verb phrase} (e.g., "Mary smokes cigarettes"). 
Learning about the behaviors of others is useful because they tell us about what that person is like more generally [e.g., @Repacholi1997; @Seiver2013].
When children describe their lives to other, a surprisingly large amount of the language produced concerns the actions of people close to them [e.g., "My brother works part-time at the restaurant"; @McGuire1986].
Habitual language about people may be a more conservative form of *trait language* (e.g., "Mary is a smoker.") and convey that behaviors are relatively enduring [@Gelman1999; @Gelman2004].

To test the generality of our theory, we use the same computational model and follow the same general experimental structure as in the first case study. 
We take event analogue of prevalence to be the *rate* with which the event occurs (e.g., how often Mary smokes).
For the generalization "Mary smokes", instances of the category being generalized are *instances of Mary* and if for the instance to have the feature, it means the instance of Mary is one in which she is smoking.
We test this by first measuring the prevalence (rate) prior distribution for various actions (e.g., how often different people smoke cigarettes; Expt. 2a).
We then measure endorsements of habitual statements while manipulating the *referent-prevalence*  (Expt. 2b), and use our computational model to predict habitual endorsements. 
<!-- We estimate the priors using a novel data-analytic approach that takes advantage of the mixture-model validated in Expt. 1b. -->
By describing novel characters to participants, we are able to directly the manipulate the *referent-prevalence* (rate), which we were unable to do for generalizations about familiar categories.
<!-- In addition to directly manipulating the referent feature-probability, these experiments provide the first test of the generality of our theory: It can apply to generalizations about events as well as categories. -->

Finally, if habituals (and generics) are truly language for conveying generalizations, it would be useful for them to reflect speakers' expectations, not merely their observations. 
Our computational model predicts endorsement rates given a referent-prevalence $p_{fk}$. 
Does prevalence represent the actual, past frequency of the event (e.g., how often Mary has actually smoked in the past month) or the predictive probability of the event in the near future (e.g., how often we should expect Mary to smoke next week)?
We extend our theory by measuring participants' predictions about future rates and endorsements of habituals when causal forces intervene on the world (Expt. 2c).
We then use our model to test whether habitual endorsements are based on past frequencies or future expectations.

@@

In order to generate model predictions for corresponding habitual statement, we first elicit the prior distributions over rates for different events.
For language about the behaviors of people, $P(p_f)$ represents a language user's background knowledge about the rates with which people do a behavior; this prior can be constructed as a distribution over *different people*, each of whom do the behavior with a different rate.
We designed our elicitation task to take advantage of the mixture-model representation of the prevalence prior used in Expt. 1 .
In particular, we assume, to a first approximation, that the distribution over prevalence can be represented as a mixture of those who tend to do the action with a stable rate and those who do not.^[
  In the case of event knowledge about people, we expect there to be several more than just these two possibilities represented in the prior, corresponding to individuals with different traits or demographics.
  We assume here a simple structure so as to not make the specification of the prior overly complex.
]
With the further assumption that, all else being equal, past is predictive of future behavior, we operationalize these two kinds of people as *people who have done the action before* and *people who have not the action before*.
We design this experiment to measure participants' beliefs about the relative proportion of these two kinds of people (as a measure of the mixture parameter in the model) as well as the rate at which people (who have done the action before) do the action (as a measure of the mean of stable-rate people).
We will assume for simplicity that people who have never done the action before will rarely or never do the action.

@@

```{r habituals-priors-subfigs, fig.width = 8}
example.habituals <- c(#"smokes marijuana", 
                       "smokes cigarettes","drinks coffee",
                       "climbs mountains", "hikes", "runs", 
                       #"wears a suit", "wears a watch", "wears socks", 
                       #"writes poems", 
                       "goes to the movies"#, "plays the banjo"
                       )
annualRates = list("5 years" = 1/5, "2 years" = 1/2,
                   "year" = 1, "6 months" = 2, "2 months" = 6,
                   "month" = 12, "2 weeks" = 26 ,"week" = 52)

@@

figure.habituals.priors.scatter <- ggplot(rs.habituals.bda.posterior.predictive.measurementSpace,
       aes(x = MAP, xmin = cred_lower, xmax = cred_upper,
           y = empirical_mean, ymin = ci_lower, ymax = ci_upper, color = gender))+
  facet_wrap(~measurement, scales = 'free', nrow = 2)+
  scale_color_solarized()+
  geom_point()+
  geom_errorbar(alpha = 0.3)+geom_errorbarh(alpha = 0.3)+
  geom_label_repel(data = rs.habituals.bda.posterior.predictive.measurementSpace %>%
                    filter(item %in% example.habituals, gender == "female"), 
                  inherit.aes =F, segment.alpha = 0.1,label.size = 0.15,
                  aes(x = MAP,y = empirical_mean, label = item), force = 15, size = 2)+
  coord_fixed()+
  xlab("Model posterior predictive")+
  ylab("Empirical means")+
  theme(legend.position = c(0.86, 0.09))

figure.habituals.priors.marginals <- ggplot(habituals.priors.marginals %>%
  filter(val > -4.9) %>%
  mutate(gender = factor(gender, levels = c("female", "male"))), 
  aes( x = val, color = gender, lty = src))+
  geom_density(size = 0.4, aes( y = ..scaled.. ))+
  facet_wrap(~item, nrow = 4)+
  geom_rect(data = rs.habituals.bda.posterior.predictive.measurementSpace %>%
               filter(gender == 'male', measurement == "% people who have done action before"), 
            aes(xmin = -5, xmax = -4.5, ymin = 0, ymax = 1 - MAP),
              fill = "#268bd2", 
            color = 'black',
             inherit.aes = F)+
  geom_rect(data = rs.habituals.bda.posterior.predictive.measurementSpace %>%
               filter(gender == 'female', measurement == "% people who have done action before"), 
            aes(xmin = -4.25, xmax = -3.75,
                                            ymin = 0, ymax = 1 - MAP),
            fill = "#cb4b16",
            color = 'black',
             inherit.aes = F)+
  scale_color_solarized()+
  xlab("Frequency")+
  ylab("Scaled probability density")+
  scale_y_continuous(limits = c(0,1), breaks = c(0, 1)) +
  theme(legend.position = c(0.93, -0.1),
        legend.title = element_blank(),
        legend.box = 'vertical',
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
        legend.margin=margin(t=0, r=0, b=-0.25, l=0, unit="cm"),
    strip.text = element_text(size = 7))+
  scale_x_continuous(
      limits = c(-5, 8),
      breaks = c(-5, 0, 2.5, 4, 5.9),
      labels = c("almost never", "annually", "monthly", "weekly", "daily")
      )+
  guides(color = F)
```

>>>>

As with categories, particular events like "Mary smoked yesterday" can be generalized into *habitual sentences*: "Mary smokes". 
<!-- Events are a broad class of entities.  -->
In our second case study, we focus on habituals about people's behaviors that take the form: \textsc{singular noun phrase} $+$ \textsc{present tense simple verb phrase} (e.g., "Mary smokes cigarettes"). 
Learning about the behaviors of others is useful because they tell us about what that person is like more generally [e.g., @Repacholi1997; @Seiver2013].
When children describe their lives to other, a surprisingly large amount of the language produced concerns the actions of people close to them [e.g., "My brother works part-time at the restaurant"; @McGuire1986].
Habitual language about people may be a more conservative form of *trait language* (e.g., "Mary is a smoker.") and convey that behaviors are relatively enduring [@Gelman1999; @Gelman2004].

To test the generality of our theory, we use the same computational model and follow the same general experimental structure as in the first case study. 
We take event analogue of prevalence to be the *rate* with which the event occurs (e.g., how often Mary smokes).^[
  Specifically, for the generalization "Mary smokes", the instances being generalized are *instances of Mary*, and for the instance to have the feature, it means the instance of Mary is one in which she is smoking.
]
We test this by first measuring the prevalence (rate) prior distribution for various actions (e.g., how often different people smoke cigarettes; Expt. 2a).
We then measure endorsements of habitual statements while manipulating the referent prevalence  (Expt. 2b), and use our computational model to predict habitual endorsements. 
<!-- We estimate the priors using a novel data-analytic approach that takes advantage of the mixture-model validated in Expt. 1b. -->
By describing novel characters to participants, we are able to directly the manipulate the referent prevalence, which we were unable to do for generalizations about familiar categories.
<!-- In addition to directly manipulating the referent feature-probability, these experiments provide the first test of the generality of our theory: It can apply to generalizations about events as well as categories. -->

Finally, if habituals (and generics) are truly language for conveying generalizations, they should reflect speakers' expectations, not merely their observations. 
Our computational model predicts endorsement rates given a referent prevalence $p$. 
Does prevalence represent the actual, objective frequency in the world or the subjective, predictive belief in the head?
We elucidate this aspect of our theory by measuring participants' predictions about future rates and endorsements of habituals when causal forces intervene on the world (Expt. 2c).
We then use our model to test whether habitual endorsements are based on past frequencies or future expectations.

<!-- (e.g., how often Mary has actually smoked in the past month) or the predictive probability of the event in the near future (e.g., how often we should expect Mary to smoke next week)? -->

@@

In order to generate model predictions for corresponding habitual statement, we first elicit the prior distributions over rates for different events.
For language about the behaviors of people, $P(p)$ represents a language user's background knowledge about the rates with which people do a behavior; this prior can be constructed as a distribution over *different people*, each of whom do the behavior with a different rate.
We designed our elicitation task to take advantage of the mixture-model representation of the prevalence prior used in Expt. 1 .
In particular, we assume, to a first approximation, that the distribution over prevalence can be represented as a mixture of those who tend to do the action with a stable rate and those who do not.^[
  In the case of event knowledge about people, we expect there to be several more than just these two possibilities represented in the prior, corresponding to individuals with different traits or demographics.
  We assume here a simple structure so as to not make the specification of the prior overly complex.
]
With the further assumption that, all else being equal, past is predictive of future behavior, we operationalize these two kinds of people as *people who have done the action before* and *people who have not the action before*.
We design this experiment to measure participants' beliefs about the relative proportion of these two kinds of people (as a measure of the mixture parameter in the model) as well as the rate at which people (who have done the action before) do the action (as a measure of the mean of stable-rate people).
We will assume for simplicity that people who have never done the action before will rarely or never do the action.

@@

```{r habituals-priors-subfigs, fig.width = 8}
example.habituals <- c(#"smokes marijuana", 
                       "smokes cigarettes","drinks coffee",
                       "climbs mountains", "hikes", "runs", 
                       #"wears a suit", "wears a watch", "wears socks", 
                       "writes poems", 
                       "writes novels", 
                       "goes to the movies", "plays the banjo"
                       )
annualRates = list("5 years" = 1/5, "2 years" = 1/2,
                   "year" = 1, "6 months" = 2, "2 months" = 6,
                   "month" = 12, "2 weeks" = 26 ,"week" = 52)

@@

figure.habituals.priors.scatter <- 
  rs.habituals.bda.posterior.predictive.measurementSpace %>%
  select(-mean, -ci_lower, -ci_upper, -empirical_mean, -n, -cred_upper, -cred_lower) %>%
  spread(measurement, MAP) %>%
  ggplot(.,
       aes(x = `% people who have done action before`, 
           y = `Frequency of doing action (log scale)`, 
           color = gender))+
  scale_color_solarized()+
  geom_point()+
  #geom_errorbar(alpha = 0.3)+geom_errorbarh(alpha = 0.3)+
  geom_label_repel(data = . %>%
                    filter(item %in% example.habituals, gender == "female"), 
                  inherit.aes =F, segment.alpha = 1,label.size = 0.25,
                  segment.size = 5,
                  aes(x = `% people who have done action before`, 
           y = `Frequency of doing action (log scale)`, 
           label = item), force = 1, size = 2)+
  coord_fixed(1/7.5)+
  #xlab("Model posterior predictive")+
  #ylab("Empirical means")+
  theme(legend.position = c(0.83, 0.16))+
  scale_y_continuous(
      limits = c(-1.5, 6),
      breaks = c(0, 2.5, 4, 5.9),
      labels = c( "annually", "monthly", "weekly", "daily")
    )+
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1))

# figure.habituals.priors.scatter <- ggplot(rs.habituals.bda.posterior.predictive.measurementSpace,
#        aes(x = MAP, xmin = cred_lower, xmax = cred_upper,
#            y = empirical_mean, ymin = ci_lower, ymax = ci_upper, color = gender))+
#   facet_wrap(~measurement, scales = 'free', nrow = 2)+
#   scale_color_solarized()+
#   geom_point()+
#   geom_errorbar(alpha = 0.3)+geom_errorbarh(alpha = 0.3)+
#   geom_label_repel(data = rs.habituals.bda.posterior.predictive.measurementSpace %>%
#                     filter(item %in% example.habituals, gender == "female"), 
#                   inherit.aes =F, segment.alpha = 0.1,label.size = 0.15,
#                   aes(x = MAP,y = empirical_mean, label = item), force = 15, size = 2)+
#   coord_fixed()+
#   xlab("Model posterior predictive")+
#   ylab("Empirical means")+
#   theme(legend.position = c(0.86, 0.09))

figure.habituals.priors.marginals <- ggplot(habituals.priors.marginals %>%
  filter(val > -4.9) %>%
  mutate(gender = factor(gender, levels = c("female", "male"))), 
  aes( x = val, color = gender, lty = src))+
  geom_density(size = 0.4, aes( y = ..scaled.. ))+
  facet_wrap(~item, nrow = 4)+
  geom_rect(data = rs.habituals.bda.posterior.predictive.measurementSpace %>%
               filter(gender == 'male', measurement == "% people who have done action before"), 
            aes(xmin = -5, xmax = -4.5, ymin = 0, ymax = 1 - MAP),
              fill = "#268bd2", 
            color = 'black',
             inherit.aes = F)+
  geom_rect(data = rs.habituals.bda.posterior.predictive.measurementSpace %>%
               filter(gender == 'female', measurement == "% people who have done action before"), 
            aes(xmin = -4.25, xmax = -3.75,
                                            ymin = 0, ymax = 1 - MAP),
            fill = "#cb4b16",
            color = 'black',
             inherit.aes = F)+
  scale_color_solarized()+
  scale_linetype_manual(values = c(2, 1))+
  xlab("Frequency")+
  ylab("Scaled probability density")+
  scale_y_continuous(limits = c(0,1), breaks = c(0, 1)) +
  theme(legend.position = c(0.93, -0.1),
        legend.title = element_blank(),
        legend.box = 'vertical',
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1),
        legend.margin=margin(t=0, r=0, b=-0.25, l=0, unit="cm"),
    strip.text = element_text(size = 7))+
  scale_x_continuous(
      limits = c(-5, 8),
      breaks = c(-5, 0, 2.5, 4, 5.9),
      labels = c("almost never", "annually", "monthly", "weekly", "daily")
      )+
  guides(color = F)
```
